Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-08 paper.tex
edits to 1 and 2.  tightened up the summary in the intro.  and the cooling time discusion also improved with outer stability limit now mentioned.

A purely hydrodynamic mechanism could circumvent difficulties with non-ideal MHD, but must overcome
the strong centrifugal stability imposed by the positive radial specific angular
momentum gradient in nearly Keplerian disks \citep{balbus96}.
One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI, \citealp{urpin98, urpin03, nelson13}, hereafter \citetalias{nelson13}).  The
basic mechanism of the VSI  in disks was first 
identified in the context of differentially rotating stars (\citealp{goldreich67}, hereafter \citetalias{goldreich67}, \citealp{fricke68}).   
The VSI arises when vertical shear, i.e.\ a variation in orbital motion along the rotation axis, 
destabilizes inertial-gravity waves, which are oscillations 
with rotation and buoyancy as restoring forces. Vertical shear
occurs wherever the disk is baroclinic, i.e.\ when constant 
density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
practically unavoidable in astrophysical disks, except at special locations like the
midplane.

To overcome centrifugal stabilization, the VSI triggers motions which are vertically elongated 
and radially narrow.  %This geometry minimizes the effects of the radial gradient of angular momentum.
Vertical elongation taps the free energy of the vertical shear \citep{umurhan13}, but is also subject to 
the stabilizing effects of vertical buoyancy if the disk is stably
stratified. To overcome vertical buoyancy, the VSI requires a short cooling time
(\citetalias{goldreich67}; \citetalias{nelson13}). Rapid radiative 
cooling, equivalently a short thermal relaxation timescale, brings a
moving fluid element into thermal equilibrium with 
its surroundings, thereby diminishing buoyancy. %not obvious actually
                                %- requires assumption of ideal gas
                                %e.o.s. and pressure eqm established
                                %quickly 
An isothermal equation of state implies instantaneous cooling (and
heating), and is the ideal context for studying the VSI
(\citealp{urpin03}; \citetalias{nelson13}; \citealp{mcnally14}, hereafter \citetalias{mcnally14}; \citealp{barker15},
hereafter \citetalias{barker15}).

Alternatively, vertical buoyancy can be eliminated by strong internal
heating, i.e.\ by the onset of convection, so that the disk becomes 
neutrally stratified in the vertical direction
\citepalias{nelson13,barker15}. However, realistic PPDs should be vertically
stably stratified in the outer regions, beyond $\sim 1$--5 AU, where
heating is  irradiation dominated \citep{bitsch15}.  In the inner
disk, strong vertical buoyancy is possible if accretion heating is
weak or is concentrated in surface layers \citep{gammie96,lesniak11}.

A full understanding of the VSI in real disks therefore requires
considering finite, non-zero cooling times, $t_{\rm c}$. Non-linear
hydrodynamical simulations with a prescribed  $t_{\rm c}$ find that VSI
turbulence in stably stratified disks requires rapid cooling  
with $t_{\rm c} \lesssim 0.01P_{\mathrm{orb}}$, where $P_\mathrm{orb}$ 
%=2\pi/\Omega$ 
is the local orbital period %and $\Omega$ is the orbital
%frequency 
 \citepalias{nelson13}. \citetalias{nelson13} further find
that the VSI can drive moderately strong transport, with Reynolds
stresses of $\alpha \sim 10^{-3}$ times the mean pressure. Simulations
with realistic radiative transfer, in lieu of a parameterized $t_{\rm
  c}$, find that the VSI in irradiated disks drive transport with
$\alpha \sim 10^{-4}$  in a  $\sim 2$ -- 10AU PPD model \citep{stoll14}.

To reach hydrodynamic turbulence via the VSI, a disk should  
first develop the linear instability. A quantitative 
examination of the thermodynamic requirement for the linear VSI is
thus essential in assessing its 
relevance to PPDs. In this work, we present a
comprehensive study of the linear VSI in vertically global disk models 
with non-zero cooling timescales, which permit buoyancy forces. 
% In this work, we present a comprehensive study of the linear VSI in vertically global disk models 
% with non-zero cooling timescales $t_c > 0$, which permit buoyancy forces. 
% Our effort generalizes both early analyses of the VSI in vertically
% localized disk models \citep{urpin98,urpin03} and recent calculations in vertically global 
% disks \citepalias{nelson13,mcnally14,barker15}. These works either
% consider the limit of instantaneous cooling ($t_c\to0$) or neutrally
% stratified disks, for which buoyancy forces play no role. 
Our effort generalizes previous studies of the linear VSI in 
vertically local \citep{urpin98,urpin03} and vertically global 
\citepalias{nelson13,mcnally14,barker15} disk models that consider the limit
of instantaneous cooling, for which buoyancy plays no role.  
%to study of how finite cooling times, and thus buoyancy, 
%affect the VSI. Our effort may also be regarded
%as a extension of early analyses of the VSI in 
%vertically localized disk models with cooling \citep{urpin98,urpin03} to
%vertically global disks with cooling. 
We consider disk  
models with vertically constant $t_{\rm c}$, as well as more realistic
cooling timescales based on dust opacity in both the optically thick
(diffusive) and thin regimes, where appropriate. Coupled with the 
results from numerical simulations, our results can be used to 
understand how the VSI operates over a wide range of model PPDs.

%\emph{TODO: the following is still a bit too wordy IMO but we can redo as a final step after these sections are finalized.}

This paper is organized as follows. In \S\ref{vsi_require} we 
introduce and qualitatively explain our main result of a critical or
characteristic cooling timescale, beyond which the VSI is either
suppressed or inefficient. We then present the  
formal linear problem in \S\ref{setup}---\ref{numerical}.    
In \S\ref{setup} we list the governing equations and global
equilibrium disk models under consideration, and formulate the linear
stability problem in \S\ref{linear}. In 
\S\ref{analytical} we give a simplified analytical discussion for
vertically isothermal disks, where we derive the aforementioned
characteristic upper limit on the cooling or thermal timescale for the VSI. 
% Our treatment is grossly simplified owing to a
% number of approximations made, but closely parallels previous work, 
% allowing us to highlight the essential effects of vertical shear, and to
% derive a characteristic upper limit to the thermal timescale for the
% VSI. This timescale is \emph{much shorter} than the dynamical
% timescale. 
In \S\ref{numerical} we numerically
solve the linear problem, and  
find basic agreement with our analysis.
% , including
% the aforementioned characteristic thermal timescale. 
In 
\S\ref{application} we apply our results to the Minimum Mass
Solar Nebula (MMSN) by modeling cooling timescales based on realistic
disk properties, and find that the VSI can operate effectively in the
outer parts of the MMSN. We summarize in \S\ref{summary} with a
discussion of caveats and possible directions for future work.

@@

\section{Why must cooling be so fast?}\label{vsi_require}     
In a stably stratified thin disk, the VSI  
requires a thermal timescale $t_c$ much shorter than the
disk dynamical timescale, 
\begin{align}
  t _c \ll \OmK^{-1}. 
\end{align}
where $\OmK$ is the Keplerian frequency (N13).    
This requirement --- that the cooling time be much shorter than an orbital period, 
which in turn is much smaller than the relevant oscillation period of vertically elongated gravity waves  --- is quite stringent.
It highlights the fact that vertical buoyancy is strongly 
stabilizing. Rapid radiative damping is thus required to weaken
buoyancy and allow the weak vertical shear to drive instability.

@@

For a thin disk ($h\ll 1$), the vertical shear rate varies with height, $z$, from the midplane as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega }{\p z} \simeq  h q \OmK \left(\frac{z}{2 H}\right),
\end{align}
where $H=hr$ is the characteristic disk scale height. 
%for several disk scale heights, $H = hr$. 
%As noted in
%\S\ref{intro} this vertical variation is a source of free energy and
%thus vertical shear ($\p_z\Omega \neq0$) is destabilizing. 
%
%For $\gamma>1$ the disk is stably stratified with a real 
This destabilizing shear competes with the stabilizing vertical
buoyancy frequency $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \OmK^2.  
\end{align}
Fig. \ref{eqm_structure} maps the vertical shear rate and buoyancy
frequency (as well as the gas density) 
in a fiducial PPD model considered in \S\ref{application}. 
%, whose parameters are in Appendix \ref{mmsn} . %Here $\rho_0$
%is the midplane density and vertical shear rate and buoyancy
%frequencies are normalized by the local Keplerian frequency. 
Vertical shear is 
generally weak compared to buoyancy, suggesting stability. In fact, 
without any cooling, the Solberg-Hoiland criteria \citep{tassoul78}  
confirms that vertical buoyancy is strong
enough to ensure (axisymmetric) stability, see \S\ref{solberg}.  With radiative cooling, 
thermal fluctuations decay which, combined with pressure equilibration, reduces the effective buoyancy.

@@

How short must $t_c$ be for vertical shear to prevail?  We start by 
considering the Richardson number $\mathrm{Ri} = N_z^2/(r\p_z\Omega)^2$, a 
ratio of buoyant to shear energies.  Though not precisely applicable to our problem,
non-rotating shear flows are stable if $\mathrm{Ri} > 1/4$ (\citealp{chandrasekhar61}, 
also see \citealp{ys02, lee10} for applications to thin dust layers in PPDs).
Following \cite{urpin03} and \cite{townsend58},  we reduce the buoyant energy by the 
ratio of cooling to forcing timescales, $t_c |r \p_z\Omega| \leq 1$ (with no reduction in buoyant energy 
expected when this inequality fails).  With this reduction, the Richardson-like criterion for 
shear \emph{instability} becomes
\begin{align}\label{bcrit_est}
  t_c \lesssim \frac{\left|r\p\Omega/\p z\right|} {N_z^2}, 
\end{align}
which is Eq.  47 of \cite{urpin03}. %This result is far from rigorous, but it contains some useful insights.
If we interpret Eq.\ \ref{bcrit_est} as a local criterion, then we see that 
the maximum cooling time which permits instability decreases with height as $1/|z|$;  
the stabilizing effect of vertical buoyancy increases more rapidly away from
the midplane than the destabilizing effect of vertical shear.  
This finding is relevant to the radiative damping of so called `surface 
modes', which we consider in \S\ref{numerical}.

For vertically global disturbances considered in this work, 
% To roughly obtain a global cooling time criterion, we evaluate Eq.\ \ref{bcrit_est}
% at $z \sim H$.  To get the precise form we want, we choose $z = \gamma H / 2$ to 
we obtain the cooling requirement 
\begin{align}\label{prelim_bcrit}
  t_c <  h |q| \OmK^{-1}/ (\gamma - 1)
\end{align}   
for the VSI to operate effectively in a vertically isothermal, thin
disk. We derive this result more formally in
\S\ref{iso_vsi_beta_crit}, but notice that Eq. \ref{prelim_bcrit}
may be obtained from Eq. \ref{bcrit_est} by evaulating it at the
height $|z|=\gamma H/2$. %i.e., at $z=O(H)$

@@

We now consider whether, and where, thermal timescales in realistic PPDs
are short enough to support VSI growth.  Fig. \ref{bcrit_mmsn2d} plots the
 thermal timescales in our fiducial PPD model with a small grain dust opacity. 
 Since cooling by radiative diffusion depends on a perturbation's lengthscale, we
 plot $t_c$ for two values of the radial wavenumber,  $k_x$,  the relevant 
 component for VSI modes that are radially narrow.  The red line
marks the boundary implied by Eq. \ref{prelim_bcrit}.  We caution that 
 Eq. \ref{prelim_bcrit} applies when $t_c$ is vertically uniform.  This section is 
 aimed at basic insights; while more realistic spatial variations in
 $t_c$ is considered in \S\ref{application}. 
 %analyzes the VSI with realistic spatial variations in.

 Due to the smaller values of $t_c$ in the outer disk, we generally expect the VSI to 
 operate outside $\sim 10$ AU.  This conclusion changes however, if
 the dust opacity is reduced,  
slowing down cooling, as we show in \S\ref{application}. % (e.g.\ Fig.\
% \ref{mmsn_bcrit_bcool}). 
In the inner disk, thermal times are long in the midplane, but short at higher altitutdes. 
Again, this structure necessitates the more detailed analyses that follow.  Nevertheless, we can already 
see that due to radiative diffusion higher wavenumbers can cool more rapidly.  Thus smaller scale disturbances 
may trigger VSI at smaller disk radii.

On the other hand, one cannot obtain arbitrarily short cooling times
by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
\citetalias{barker15}), because very small wavelengths are in the
optically thin cooling regime,  where cooling is independent of
lengthscale. Furthermore, 
%while we do not include viscosity in our
%idealized disk models, in reality 
small-scale perturbations will be 
prone to viscous damping, including the effective viscosity from turbulent motions.
While our linear, inviscid calculations do not address this issue directly, we expect that 
shorter wavelength modes will drive weaker non-linear (and non-axisymmetric) transport.

% However, we cannot and do not find that VSI will always be relevant for
% sufficiently small perturbations because \begin{inparaenum}[(i)] \item very small wavelengths are in the optically 
% thin cooling regime,  where cooling is independent of lengthscale;
% and \item small scale perturbations 
% are more prone to viscous damping, and should drive weaker transport.
% \end{inparaenum}

We now proceed with our detailed analysis of the VSI to arrive
at Eq. \ref{prelim_bcrit} and verify it numerically. The interested 
reader may jump to \S\ref{application} for the application of our
results to PPDs.

>>>>

A purely hydrodynamic mechanism could circumvent difficulties with non-ideal MHD, but must overcome
the strong centrifugal stability imposed by the positive radial specific angular
momentum gradient in nearly Keplerian disks \citep{balbus96}.
One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI, \citealp{urpin98, urpin03, nelson13}, hereafter \citetalias{nelson13}).  The
basic mechanism of the VSI  in disks was first 
identified in the context of differentially rotating stars (\citealp{goldreich67}, hereafter \citetalias{goldreich67}, \citealp{fricke68}).   
The VSI arises when vertical shear, i.e.\ a variation in orbital motion along the rotation axis, 
destabilizes inertial-gravity waves, which are oscillations 
with rotation and buoyancy as restoring forces. 
% \citep{lubow93}. 
Vertical shear
occurs wherever the disk is baroclinic, i.e.\ when constant 
density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
practically unavoidable in astrophysical disks, except at special locations like the
midplane.

To overcome centrifugal stabilization, the VSI triggers motions which are vertically elongated 
and radially narrow.  %This geometry minimizes the effects of the radial gradient of angular momentum.
Vertical elongation taps the free energy of the vertical shear \citep{umurhan13}, but is also subject to 
the stabilizing effects of vertical buoyancy if the disk is stably
stratified. To overcome vertical buoyancy, the VSI requires a short cooling time
(\citetalias{goldreich67}; \citetalias{nelson13}). Rapid radiative 
cooling, i.e.\ a short thermal relaxation timescale, brings a
moving fluid element into thermal equilibrium with 
its surroundings, thereby diminishing buoyancy. %not obvious actually
                                %- requires assumption of ideal gas
                                %e.o.s. and pressure eqm established
                                %quickly 
An isothermal equation of state implies instantaneous thermal relaxation, and is the ideal context for studying the VSI
(\citealp{urpin03}; \citetalias{nelson13}; \citealp{mcnally14}, hereafter \citetalias{mcnally14}; \citealp{barker15},
hereafter \citetalias{barker15}).

Alternatively, vertical buoyancy can be eliminated by strong internal
heating, i.e.\ by the onset of convection, so that the disk becomes 
neutrally stratified in the vertical direction
\citepalias{nelson13,barker15}. However, realistic PPDs should be vertically
stably stratified in the outer regions, beyond $\sim 1$--5 AU, where
heating is  irradiation dominated \citep{bitsch15}.  Even in the inner
disk, strong vertical buoyancy is possible if accretion heating is
weak or is concentrated in surface layers \citep{gammie96,lesniak11}.

Understanding the VSI in real disks therefore requires
considering finite, non-zero cooling times, $t_{\rm c}$. Non-linear
hydrodynamical simulations with a prescribed  $t_{\rm c}$ find that VSI
turbulence in stably stratified disks requires rapid cooling  
with $t_{\rm c}$ shorter than orbital timescales
 \citepalias{nelson13}. When the cooling time is short enough, 
 the VSI can drive moderately strong transport, with Reynolds
stresses of $\alpha \sim 10^{-3}$ times the mean pressure \citepalias{nelson13}. Simulations
with realistic radiative transfer, in lieu of a fixed $t_{\rm
  c}$, find that the VSI in irradiated disks drive transport with
$\alpha \sim 10^{-4}$  in a  $\sim 2$ -- 10AU PPD model \citep{stoll14}.

Studying the linear growth of the VSI is necessary for 
understanding how it can ultimately drive turbulent transport.
The pioneering linear analyses of the VSI considered vertically 
local disturbances \citep{urpin98,urpin03}.  A vertical global analysis 
(e.g.\ \citetalias{nelson13,mcnally14,barker15}) 
is essential for understanding how vertically elongated disturbances 
interact with the disk's vertical structure.  Moreover a vertically global analysis
allows more direct comparisons with modern numerical simulations.  This work 
generalizes previous vertically global analyses by including a finite cooling time.

%NOTE: no need to reexplain importance of cooling or to say more about our work before the "organized as follows" summary

This paper is organized as follows. In \S\ref{vsi_require} we explain, without derivation,  
our main result for the critical cooling timescale, beyond which VSI growth is 
suppressed. %We then present the  
%formal linear problem in \S\ref{setup}---\ref{numerical}.    
 We develop our disk model in \S\ref{setup} and derive linearized
 perturbation equations in  \S\ref{linear}.
 Section \ref{analytical} contains our main analytic results, leading to 
 the derivation of the critical cooling time in \S\ref{iso_vsi_beta_crit}.
 In \S\ref{numerical} we analyze linear VSI growth, and numerically confirm the critical cooling time.
We apply our results to PPDs in \S\ref{application}, with cooling times derived from dust opacities.
We discuss caveats and extensions in \S\ref{caveats}.
 We conlude in \S\ref{summary} with a summary.  Some technical and background 
 developments are explained in the appendices.

@@

\section{Why must cooling be so fast?}\label{vsi_require}     
In a stably stratified thin disk, the VSI  
requires a thermal timescale $t_c$ significantly shorter than the
disk dynamical timescale, 
\begin{align}
  t _c \ll \OmK^{-1},
\end{align}
where $\OmK$ is the Keplerian frequency (N13).    
This requirement --- that the cooling time be much shorter than an orbital period, 
which in turn is much smaller than the relevant oscillation period of vertically elongated gravity waves  --- is quite stringent.
It highlights the fact that vertical buoyancy is strongly 
stabilizing. Rapid radiative damping is thus required to weaken
buoyancy and allow the weak vertical shear to drive instability.

@@

For a thin disk ($h\ll 1$), the vertical shear rate varies with height, $z$, from the midplane as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega }{\p z} \simeq  h q \OmK \left(\frac{z}{2 H}\right),
\end{align}
where $H=hr$ is the characteristic disk scale height. 
%for several disk scale heights, $H = hr$. 
%As noted in
%\S\ref{intro} this vertical variation is a source of free energy and
%thus vertical shear ($\p_z\Omega \neq0$) is destabilizing. 
%
%For $\gamma>1$ the disk is stably stratified with a real 
This destabilizing shear competes with the stabilizing vertical
buoyancy frequency, $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \OmK^2.  
\end{align}
Fig. \ref{eqm_structure} maps the vertical shear rate and buoyancy
frequency (as well as the gas density) 
in a fiducial PPD model, see \S\ref{toy_relax} for details. 
%, whose parameters are in Appendix \ref{mmsn} . %Here $\rho_0$
%is the midplane density and vertical shear rate and buoyancy
%frequencies are normalized by the local Keplerian frequency. 
Vertical shear is 
generally weak compared to buoyancy, suggesting stability. In fact, 
without any cooling, the Solberg-Hoiland criteria \citep{tassoul78}  
confirms that vertical buoyancy is strong
enough to ensure (axisymmetric) stability, see \S\ref{solberg}.  With radiative cooling, 
thermal fluctuations decay which, combined with pressure equilibration, reduces the effective buoyancy.

@@

How short must $t_c$ be for vertical shear to prevail?  We start with 
 the Richardson number $\mathrm{Ri} = N_z^2/(r\p_z\Omega)^2$, a 
ratio of buoyant to shear energies.  Though not precisely applicable to our problem,
non-rotating shear flows are stable if $\mathrm{Ri} > 1/4$ (\citealp{chandrasekhar61}, 
also see \citealp{ys02, lee10} for applications to thin dust layers in PPDs).
Following \cite{urpin03} and \cite{townsend58},  we reduce the buoyant energy by the 
ratio of cooling to forcing timescales, $t_c |r \p_z\Omega| \leq 1$ (with no reduction in buoyant energy 
expected when this inequality fails).  With this reduction, the Richardson-like criterion for 
shear instability becomes  \citep{urpin03}
\begin{align}\label{bcrit_est}
  t_c \lesssim \frac{\left|r\p\Omega/\p z\right|} {N_z^2}.
\end{align}
 %This result is far from rigorous, but it contains some useful insights.
If we interpret Eq.\ \ref{bcrit_est} as a local criterion, then we see that 
the maximum cooling time which permits instability decreases with height as $1/|z|$;  
the stabilizing effect of vertical buoyancy increases more rapidly away from
the midplane than the destabilizing effect of vertical shear.  
This finding is relevant to the radiative damping of so called `surface 
modes', which we consider in \S\ref{numerical}.

We are more interested, however, the the global instability criterion for disturbances at all heights.
Our desired result 
is the cooling requirement 
\begin{align}\label{prelim_bcrit}
  t_c <  h |q| \OmK^{-1}/ (\gamma - 1)
\end{align}   
for VSI growth that we derive in \S\ref{iso_vsi_beta_crit}.  We can simply 
(but non-rigourously) obtain Eq. \ref{prelim_bcrit}
  by evaulating  Eq. \ref{bcrit_est} at $|z|=\gamma H/2$. %i.e., at $z=O(H)$

@@

We now consider whether, and where, thermal timescales in realistic PPDs
are short enough to support VSI growth.  Fig. \ref{bcrit_mmsn2d} plots the
 thermal timescales in our fiducial PPD model, see \S\ref{toy_relax} for details.
 The regions between the red lines satisfy the cooling requirement of Eq.\ \ref{prelim_bcrit}.
 Even without a detailed analysis we can expect VSI growth between $\sim 10$ -- 150 AU.
 The inner disk is complicated by the fact that the optically thick midplane has longer cooling times. 
 Thus the cooling criterion fails in the midplane but is satisfied above.  This case requires the 
 more detailed analysis of \S\ref{application}.

  Fig. \ref{bcrit_mmsn2d} also compares two different radial wavenumbers, 
  $k_x = 10/H$ and $50/H$.  In the inner disk, the higher wavenumber mode cools faster, due to radiative diffusion.
  Thus in the inner disk, shorter wavelength VSI modes are more like to grow. In the outer disk, the cooling time is the 
  same for different $k_x$ and $z$, because optically thin cooling is independent of both lengthscale and density.
  Thus the outer barrier to growth, beyond $\sim 150$ AU in this model, is independent of  wavelength. 
This example highlights the fact that optically thin cooling sets a lower limit to the cooling time; a fact that is sometimes obscured when 
the radiative diffusion approximation is made at the outset.

% Since cooling by radiative diffusion depends on a perturbation's lengthscale, we
% plot $t_c$ for two values of the radial wavenumber,  $k_x$,  the relevant 
% component for VSI modes that are radially narrow.  The red line
%marks the boundary implied by Eq. \ref{prelim_bcrit}.  We caution that 
% Eq. \ref{prelim_bcrit} applies when $t_c$ is vertically uniform.  This section is 
% aimed at basic insights; while more realistic spatial variations in
% $t_c$ are considered in \S\ref{application}. 
% %analyzes the VSI with realistic spatial variations in.
% 
% Due to the smaller values of $t_c$ in the outer disk, we generally expect the VSI to 
% operate outside $\sim 10$ AU.  This conclusion changes however, if
% the dust opacity is reduced,  
%slowing down cooling, as we show in \S\ref{application}. % (e.g.\ Fig.\
%% \ref{mmsn_bcrit_bcool}). 
%In the inner disk, thermal times are long in the midplane, but short at higher altitutdes. 
%Again, this structure necessitates the more detailed analyses that follow.  Nevertheless, we can already 
%see that due to radiative diffusion higher wavenumbers can cool more rapidly.  Thus smaller scale disturbances 
%may trigger VSI at smaller disk radii. 
%
%On the other hand, one cannot obtain arbitrarily short cooling times
%by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
%\citetalias{barker15}), because very small wavelengths are in the
%optically thin cooling regime,  where cooling is independent of
%lengthscale. Furthermore, 
%%while we do not include viscosity in our
%%idealized disk models, in reality 
%small-scale perturbations will be 
%prone to viscous damping, including the effective viscosity from turbulent motions.
%While our linear, inviscid calculations do not address this issue directly, we expect that 
%shorter wavelength modes will drive weaker non-linear (and non-axisymmetric) transport. 
%
%% However, we cannot and do not find that VSI will always be relevant for
%% sufficiently small perturbations because \begin{inparaenum}[(i)] \item very small wavelengths are in the optically 
%% thin cooling regime,  where cooling is independent of lengthscale;
%% and \item small scale perturbations 
%% are more prone to viscous damping, and should drive weaker transport.
%% \end{inparaenum} 
%
%We now proceed with our detailed analysis of the VSI to arrive
%at Eq. \ref{prelim_bcrit} and verify it numerically. The interested 
%reader may jump to \S\ref{application} for the application of our
%results to PPDs.
